---
title: "Plasmid reconstruction stability versus sequencing depth"
author: ''
date: "`r Sys.Date()`"
output:
  word_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    df_print: kable
  pdf_document:
    toc: true
params:
  input: S3_plasmid_stability_scores_clean.tsv
  qc_input: plasmid_assembly_qc_exploration.tsv
  qc_exclude: !r c("7092", "7106")
  ref_depth: 60
  depths: !r c(10, 20, 40)
  # Size floor. The Methods describe the exclusion of a single small plasmid
  # (bla-OXA-232, 6.1 kb); a 10 kb floor removes the same plasmid. State
  # whichever rule the manuscript uses.
  min_size_bp: 10000
  primary_threshold: 0.05
  sensitivity_thresholds: !r c(0.03, 0.10)
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(dplyr)
library(tidyr)
# TRUE only if the package exists AND its namespace actually loads
# (a broken binary, e.g. readxl with a missing _iconv symbol, returns FALSE).
have <- function(p) isTRUE(tryCatch(requireNamespace(p, quietly = TRUE),
                                    error = function(e) FALSE))

# Read .xlsx via readxl or openxlsx (whichever is available), or a .tsv/.csv.
read_any <- function(path) {
  if (grepl("\\.xlsx?$", path, ignore.case = TRUE)) {
    if (have("readxl"))   return(as.data.frame(readxl::read_excel(path)))
    if (have("openxlsx")) return(openxlsx::read.xlsx(path))
    stop("Install 'readxl' or 'openxlsx' to read Excel, or point params$input at a .tsv")
  }
  read.delim(path, check.names = FALSE)
}
```

## Data and stability classification

Each carbapenemase-encoding plasmid was reconstructed at several downsampled
depths. Stability at a given depth is defined relative to the isolate's own
`r params$ref_depth`x assembly (the study-internal reference). A plasmid is
**stable** if its contig size falls within a set deviation of the reference *and*
the carbapenemase gene is still detected on the contig, and **unstable**
otherwise. Three failure modes are recorded separately and pooled for analysis:
size deviation beyond the threshold, loss of the carbapenemase gene, and
non-recovery of the plasmid.
Only medium-to-large plasmids (> `r params$min_size_bp/1000` kb) are analysed;
the single small plasmid is considered separately.

Sizes are parsed by taking the leading number of each cell, after removing any
parenthetical annotation and repairing a dot or comma used as a thousands
separator. Annotations recording that the carbapenemase gene was not recovered
at a given depth are read separately and make that plasmid unstable at that
depth regardless of contig size.

```{r load}
parse_size <- function(x) {
  # Take the leading number only. Annotations in the source table may themselves
  # contain digits (e.g. "47097 (missing KPC-3)"), so stripping all non-digits
  # would absorb them and inflate the size.
  x <- trimws(format(x, scientific = FALSE, trim = TRUE))
  x <- sub("\\s*\\(.*$", "", x)                                   # drop "( ... )" annotation
  x <- gsub("[.,](?=[0-9]{3}\\b)", "", x, perl = TRUE)            # 169.978 -> 169978
  x <- sub("^([0-9]+).*$", "\\1", x)                               # leading digits only
  suppressWarnings(as.numeric(ifelse(grepl("^[0-9]+$", x), x, NA)))
}

# TRUE where the annotation in a size cell says the carbapenemase gene was not
# recovered at that depth. Such plasmids are unstable regardless of contig size.
gene_absent <- function(x) {
  x <- tolower(as.character(x))
  !is.na(x) & (grepl("missing", x) | grepl("no carbapenemase", x))
}

ref_col <- paste0("size_", params$ref_depth, "x")
dep_cols <- paste0("size_", params$depths, "x")

raw <- read_any(params$input)
stopifnot(all(c(ref_col, dep_cols) %in% names(raw)))

dat <- raw %>%
  mutate(across(all_of(c(ref_col, dep_cols)), parse_size, .names = "{.col}_n")) %>%
  mutate(across(all_of(dep_cols), gene_absent, .names = "{.col}_noGene")) %>%
  mutate(ref_n = .data[[paste0(ref_col, "_n")]])

# Drop footer / caption / blank rows: keep only rows identifying a real plasmid
id_cols <- intersect(c("gene_symbol", "replicon_type"), names(dat))
if (length(id_cols) > 0) {
  is_real <- Reduce(`&`, lapply(id_cols, function(cc)
    !is.na(dat[[cc]]) & trimws(as.character(dat[[cc]])) != ""))
  dat <- dat[is_real, , drop = FALSE]
}

dat <- dat %>%
  mutate(plasmid_id = dplyr::row_number()) %>%
  filter(!is.na(ref_n), ref_n > params$min_size_bp)

cat(sprintf("Plasmids > %.0f kb with a %dx reference: %d\n",
            params$min_size_bp/1000, params$ref_depth, nrow(dat)))

# long format: one row per plasmid x depth, with a stability call for a threshold
to_long <- function(df, thr) {
  sizes <- df %>%
    select(plasmid_id, isolate, gene_symbol, replicon_type, ref_n,
           all_of(paste0(dep_cols, "_n"))) %>%
    pivot_longer(all_of(paste0(dep_cols, "_n")),
                 names_to = "depth", values_to = "size") %>%
    mutate(depth = as.integer(gsub("\\D", "", depth)))
  genes <- df %>%
    select(plasmid_id, all_of(paste0(dep_cols, "_noGene"))) %>%
    pivot_longer(-plasmid_id, names_to = "depth", values_to = "no_gene") %>%
    mutate(depth = as.integer(gsub("\\D", "", depth)))
  sizes %>%
    left_join(genes, by = c("plasmid_id", "depth")) %>%
    mutate(no_gene = tidyr::replace_na(no_gene, FALSE),
           within_size = !is.na(size) & abs(size - ref_n) / ref_n <= thr,
           stable = as.integer(within_size & !no_gene),
           failure = dplyr::case_when(
             stable == 1L      ~ "stable",
             is.na(size)       ~ "plasmid not recovered",
             no_gene           ~ "carbapenemase gene not detected",
             TRUE              ~ "size deviation > threshold"))
}
```

## Stable counts by depth (all thresholds)

```{r counts}
all_thr <- sort(unique(c(params$primary_threshold, params$sensitivity_thresholds)))
counts <- lapply(all_thr, function(thr) {
  to_long(dat, thr) %>%
    group_by(depth) %>%
    summarise(n = n(), n_stable = sum(stable),
              pct_stable = round(100 * mean(stable), 1), .groups = "drop") %>%
    mutate(threshold = paste0(thr * 100, "%"))
}) %>% bind_rows()
tidyr::pivot_wider(counts, names_from = depth,
                   values_from = c(n_stable, pct_stable, n))
```

## Failure modes at the primary threshold

Reported separately so that size-driven and gene-driven failures can be
distinguished, then pooled for the models below.

```{r failure-modes}
to_long(dat, params$primary_threshold) %>%
  count(depth, failure) %>%
  tidyr::pivot_wider(names_from = failure, values_from = n, values_fill = 0)
```

## Reference (60x) assembly quality

The 60x reference assemblies underpinning the stability analysis were assessed
for reconstruction reliability. Flags mark carbapenemase-encoding contigs that
are non-circular, contain repeats, have high multiplicity (> 1.5x), or have a
high plasmid-to-chromosome coverage ratio (> 1.5x). These are reported as a
reference-reliability caveat and were interpreted with caution; they do not
define per-depth stability.

```{r qc}
qc <- read.delim(params$qc_input, check.names = FALSE)
# exclude fragmented isolates (e.g. 7092) for consistency with the other analyses
if (length(params$qc_exclude) > 0)
  qc <- qc[!grepl(paste(params$qc_exclude, collapse = "|"), qc$sample), ]
flag_cols <- c("flag_non_circular", "flag_repeat",
               "flag_high_mult", "flag_high_cov_ratio")
qc$any_flag <- apply(qc[flag_cols], 1, function(r) any(r == "FLAG"))

qc_summary <- data.frame(
  criterion = c("Non-circular", "Repeat-containing", "High multiplicity (>1.5x)",
                "High coverage ratio (>1.5x)", "Any flag"),
  n_flagged = c(sum(qc$flag_non_circular == "FLAG"), sum(qc$flag_repeat == "FLAG"),
                sum(qc$flag_high_mult == "FLAG"), sum(qc$flag_high_cov_ratio == "FLAG"),
                sum(qc$any_flag)),
  total = nrow(qc)
)
qc_summary$pct <- round(100 * qc_summary$n_flagged / qc_summary$total, 1)
qc_summary

# same, restricted to the two focus replicon types
foc <- qc[qc$replicon %in% c("IncN", "IncL/M"), ]
cat(sprintf("\nIncN / IncL/M contigs: %d, of which any-flagged: %d (%.0f%%)\n",
            nrow(foc), sum(foc$any_flag), 100 * mean(foc$any_flag)))
```

## Primary analysis (`r params$primary_threshold*100`% threshold)

The same plasmids are measured at each depth, so the depth groups are not
independent. The primary test is therefore a logistic regression fitted by
generalised estimating equations (GEE), clustered by plasmid, with depth as an
ordered score. Cochran's Q is reported as a confirmatory paired omnibus test.

```{r primary}
long_p <- to_long(dat, params$primary_threshold)

# GEE logistic trend (accounts for within-plasmid correlation)
if (have("geepack")) {
  gee <- geepack::geeglm(stable ~ depth, id = plasmid_id, data = long_p,
                         family = binomial, corstr = "exchangeable")
  s <- summary(gee)$coefficients
  or <- exp(coef(gee)["depth"])
  cat(sprintf("GEE trend: OR per 1x depth = %.4f (per 10x = %.2f), p = %.3g\n",
              or, or^10, s["depth", "Pr(>|W|)"]))
} else {
  cat("Install 'geepack' for the GEE trend test.\n")
}

# Cochran's Q (paired omnibus across the three depths), base R (no packages)
wide <- long_p %>%
  select(plasmid_id, depth, stable) %>%
  tidyr::pivot_wider(names_from = depth, values_from = stable) %>%
  select(-plasmid_id) %>% as.matrix()
k  <- ncol(wide); Cj <- colSums(wide); Ri <- rowSums(wide)
Q  <- (k - 1) * (k * sum(Cj^2) - sum(Cj)^2) / (k * sum(Ri) - sum(Ri^2))
cat(sprintf("Cochran's Q = %.2f, df = %d, p = %.3g\n",
            Q, k - 1, pchisq(Q, k - 1, lower.tail = FALSE)))
```

## Comparison with the (unpaired) Cochran–Armitage test

Reported only to show the effect of ignoring the pairing; not the primary test.

The Cochran-Armitage test assumes independent groups, whereas the same plasmids
are observed at every depth. It is reported as an unclustered sensitivity check
alongside the GEE, which accounts for the repeated measures.

```{r cochran-armitage}
cochran_armitage <- function(stable, total, scores) {
  n <- sum(total); p_bar <- sum(stable) / n; scores <- as.numeric(scores)
  T_stat <- sum(scores * (stable - total * p_bar))
  V <- p_bar * (1 - p_bar) * (n * sum(scores^2 * total) - sum(scores * total)^2) / n
  z <- T_stat / sqrt(V)
  c(z = z, p_two = 2 * pnorm(-abs(z)))
}
agg <- to_long(dat, params$primary_threshold) %>%
  group_by(depth) %>% summarise(stable = sum(stable), total = n(), .groups = "drop")
round(cochran_armitage(agg$stable, agg$total, agg$depth), 4)
```

## Sensitivity analyses (`r paste(params$sensitivity_thresholds*100, collapse="% and ")`%)

```{r sensitivity}
sens <- lapply(params$sensitivity_thresholds, function(thr) {
  lp <- to_long(dat, thr)
  p_gee <- NA
  if (have("geepack")) {
    g <- geepack::geeglm(stable ~ depth, id = plasmid_id, data = lp,
                         family = binomial, corstr = "exchangeable")
    p_gee <- summary(g)$coefficients["depth", "Pr(>|W|)"]
  }
  data.frame(threshold = paste0(thr * 100, "%"),
             gee_trend_p = signif(p_gee, 3))
}) %>% bind_rows()
sens
```

## Session info

```{r session}
sessionInfo()
```
