---
title: "Sequencing depth and MLST sequence-type concordance"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: kable
params:
  # Directory holding the cleaned per-depth files, named mlst_<depth>_S1_isolates.csv
  input_dir: "."
  # "producing" = the 100 carbapenemase-producing isolates
  # "all"       = both cohorts (110 isolates)
  cohort: "producing"
  min_isolates_per_scheme: 5
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
have <- function(p) requireNamespace(p, quietly = TRUE)
```

## Reference definition

The reference ST for each isolate is the call at the **highest available
sequencing depth**, which is the value reported in Supplementary Table S1 and is
carried in the `consensus_st_S1` column of every cleaned per-depth file. This
differs from the majority-consensus-across-depths definition used previously;
the Methods text must state whichever definition is used here.

Because the reference is taken from the highest depth, it does not incorporate
the 10x call, so concordance at 10x is not measured partly against itself.

Isolates with no reference ST are untypable and are excluded, so the model
describes concordance **among typable isolates**. The outcome is coded
1 = concordant, so the model estimates P(concordant).

## Load the cleaned per-depth files

```{r load}
files <- list.files(params$input_dir, pattern = "^mlst_[0-9]+x_S1_isolates\\.csv$",
                    full.names = TRUE)
stopifnot(length(files) >= 2)

d <- do.call(rbind, lapply(files, function(f) {
  x <- read.csv(f, stringsAsFactors = FALSE, check.names = FALSE)
  req <- c("isolate_id", "cohort", "coverage", "mlst_scheme", "mlst_st",
           "consensus_st_S1", "concordant_with_S1")
  stopifnot(all(req %in% names(x)))
  x
}))

d$coverage_n <- as.integer(sub("x$", "", d$coverage))
message("Depths loaded: ", paste(sort(unique(d$coverage_n)), collapse = ", "))
```

### Analysis population

```{r population}
if (params$cohort == "producing") d <- d[d$cohort == "carbapenemase-producing", ]

# typable isolates only: a reference ST exists
d <- d[!is.na(d$consensus_st_S1) & trimws(as.character(d$consensus_st_S1)) != "", ]

# keep only isolates observed at every loaded depth, so the comparison is paired
depths <- sort(unique(d$coverage_n))
n_by_iso <- table(d$isolate_id)
complete <- names(n_by_iso)[n_by_iso == length(depths)]
dropped_incomplete <- setdiff(unique(d$isolate_id), complete)
if (length(dropped_incomplete))
  cat("Dropped for incomplete depth coverage:", paste(dropped_incomplete, collapse = ", "), "\n")
d <- d[d$isolate_id %in% complete, ]

d$concordant <- as.integer(tolower(trimws(d$concordant_with_S1)) == "yes")
stopifnot(!any(is.na(d$concordant)))

# depth predictor: 10x versus every depth at or above 20x, 10x as the baseline
d$coverage_cat <- factor(ifelse(d$coverage_n == 10, "10x", ">=20x"),
                         levels = c("10x", ">=20x"))

c(observations = nrow(d), isolates = length(unique(d$isolate_id)),
  depths_pooled_in_ge20x = sum(depths >= 20))
```

> Note on the predictor: every depth at or above 20x is pooled into one level.
> With only 10x and 20x loaded this is a paired 10x-versus-20x contrast; once
> 40x and above are added the `>=20x` arm carries more rows than the 10x arm.
> State in the Methods which depths were pooled.

## Concordance by depth

```{r concordance-by-depth}
by_cov <- aggregate(concordant ~ coverage_n, data = d, FUN = mean)
by_cov$n <- aggregate(concordant ~ coverage_n, data = d, FUN = length)$concordant
by_cov$n_concordant <- aggregate(concordant ~ coverage_n, data = d, FUN = sum)$concordant
by_cov$concordant <- round(by_cov$concordant, 3)
by_cov

plot(by_cov$coverage_n, by_cov$concordant, type = "b", pch = 19,
     xlab = "Coverage (x)", ylab = "ST concordance",
     ylim = c(min(by_cov$concordant), 1))
```

### Non-assignment versus misassignment

```{r calls-vs-mismatches}
# an ST was called (not "-") but did not match the reference
called <- trimws(d$mlst_st) != "-" & trimws(d$mlst_st) != ""
wrong  <- sum(called & d$concordant == 0)
cat("Rows where an ST was called but did NOT match the reference:", wrong, "\n")
cat("(0 means reduced concordance at low depth reflects failure to assign,",
    "not assignment of a conflicting ST)\n")
```

## Paired 10x versus 20x comparison (exact McNemar)

This is the contrast reported alongside the concordance percentages. It is
strictly paired and uses only those two depths, so it is a different comparison
from the pooled `>=20x` model below. Report both, and say so.

```{r mcnemar}
if (all(c(10, 20) %in% depths)) {
  w <- reshape(d[d$coverage_n %in% c(10, 20), c("isolate_id", "coverage_n", "concordant")],
               idvar = "isolate_id", timevar = "coverage_n", direction = "wide")
  names(w) <- c("isolate_id", "c10", "c20")
  w <- w[complete.cases(w), ]

  gained <- sum(w$c10 == 0 & w$c20 == 1)
  lost   <- sum(w$c10 == 1 & w$c20 == 0)
  cat(sprintf("Paired isolates: %d | concordant at 10x: %d (%.1f%%) | at 20x: %d (%.1f%%)\n",
              nrow(w), sum(w$c10), 100 * mean(w$c10), sum(w$c20), 100 * mean(w$c20)))
  cat(sprintf("Gained concordance 10x -> 20x: %d | lost: %d\n", gained, lost))

  bt <- binom.test(gained, gained + lost, p = 0.5)   # exact McNemar
  cat(sprintf("McNemar exact two-sided p = %.3g\n", bt$p.value))
} else {
  cat("Both 10x and 20x are required for the paired comparison.\n")
}
```

## Helper functions

```{r helpers}
or_table <- function(model, cluster) {
  if (have("sandwich") && have("lmtest")) {
    vc <- sandwich::vcovCL(model, cluster = cluster)
    ct <- lmtest::coeftest(model, vcov. = vc)
    ci <- lmtest::coefci(model, vcov. = vc)
    est <- ct[, "Estimate"]; p <- ct[, ncol(ct)]; lo <- ci[, 1]; hi <- ci[, 2]
    note <- "cluster-robust (by isolate)"
  } else {
    warning("Install 'sandwich' and 'lmtest' for clustered SEs; using model-based SEs")
    co <- summary(model)$coefficients
    est <- co[, "Estimate"]; p <- co[, ncol(co)]
    ci <- confint.default(model); lo <- ci[, 1]; hi <- ci[, 2]
    note <- "model-based (NOT clustered)"
  }
  list(tab = data.frame(term = names(est),
                        odds_ratio = round(exp(est), 3),
                        ci_lower = round(exp(lo), 3),
                        ci_upper = round(exp(hi), 3),
                        p_value = signif(p, 3), row.names = NULL),
       note = note)
}

mcfadden <- function(model, data) {
  null <- glm(concordant ~ 1, family = binomial, data = data)
  1 - as.numeric(logLik(model)) / as.numeric(logLik(null))
}
```

## Primary model: depth only, clustered by isolate

```{r primary}
m1 <- glm(concordant ~ coverage_cat, family = binomial, data = d)
r1 <- or_table(m1, cluster = d$isolate_id)
cat("SE type:", r1$note, "\n")
r1$tab
cat(sprintf("McFadden pseudo-R2 = %.3f\n", mcfadden(m1, d)))

or_cov <- r1$tab[grepl("coverage_cat", r1$tab$term), ]
cat(sprintf("\n>=20x vs 10x: OR(concordant) = %.2f (95%% CI %.2f-%.2f), p = %s\n",
            or_cov$odds_ratio, or_cov$ci_lower, or_cov$ci_upper, or_cov$p_value))
```

## Firth penalised model

Discordant calls at `>=20x` are rare, so the unpenalised fit approaches
separation and its interval is wide. Firth penalisation stabilises the estimate
and gives a profile-likelihood interval.

**Caveat to report:** `logistf` does not accept a cluster argument, so this
interval treats each isolate-depth row as independent and is narrower than the
repeated-measures design supports. Use it as a robustness check on the point
estimate and quote the clustered interval from the primary model as the primary
result, or say explicitly which one the manuscript reports.

```{r firth}
if (have("logistf")) {
  fm <- logistf::logistf(concordant ~ coverage_cat, data = d)
  data.frame(term = names(coef(fm)),
             odds_ratio = round(exp(coef(fm)), 3),
             ci_lower = round(exp(fm$ci.lower), 3),
             ci_upper = round(exp(fm$ci.upper), 3),
             p_value = signif(fm$prob, 3), row.names = NULL)
} else {
  cat("Install 'logistf' to run the Firth penalised model.\n")
}
```

## Scheme-adjusted sensitivity model

Checks that the depth effect is not driven by species composition. The scheme
used is the modal non-missing scheme across the depths loaded for each isolate,
since the cleaned files carry a per-depth scheme rather than a reference one.
Report only the depth term; scheme coefficients may be unstable.

```{r scheme-adjusted}
modal <- function(x) {
  x <- x[!is.na(x) & x != "" & x != "-"]
  if (!length(x)) return(NA_character_)
  names(sort(table(x), decreasing = TRUE))[1]
}
ref_scheme <- tapply(d$mlst_scheme, d$isolate_id, modal)
d$ref_scheme <- ref_scheme[as.character(d$isolate_id)]

cnt  <- tapply(d$isolate_id, d$ref_scheme, function(x) length(unique(x)))
keep <- names(cnt)[cnt > params$min_isolates_per_scheme]
d2   <- d[d$ref_scheme %in% keep, ]
d2$ref_scheme <- factor(d2$ref_scheme)
cat("Schemes retained:", paste(levels(d2$ref_scheme), collapse = ", "), "\n")

m2 <- glm(concordant ~ coverage_cat + ref_scheme, family = binomial, data = d2)
r2 <- or_table(m2, cluster = d2$isolate_id)
r2$tab[grepl("coverage_cat", r2$tab$term), ]
cat(sprintf("McFadden pseudo-R2 (adjusted) = %.3f\n", mcfadden(m2, d2)))
```

## Numbers to carry into the manuscript

```{r summary}
cat("Analysis population\n")
cat("  cohort scope           :", params$cohort, "\n")
cat("  typable isolates       :", length(unique(d$isolate_id)), "\n")
cat("  depths                 :", paste(sort(unique(d$coverage_n)), collapse = ", "), "\n")
cat("\nThese counts must match Supplementary Table S1. If they do not, reconcile\n")
cat("before quoting any figure in the manuscript.\n")
```

## Session info

```{r session}
sessionInfo()
```
